Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-07-17 surfaces-write-up/surface-write-up.tex
many line breaks, cleanups, relabel an eqn, add -

\begin{lem}
\label{lem:proj-relations-psi}
$\ker(\psi)$ is generated in degrees up to
\begin{align}
\label{eqn:pm-relation-degree}
	\frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} +  \sum_{i=1}^n \ell_i a_i.
\end{align}

@@

\noindent
for all $d \in \mathbb{N}$ and $c_i \ge -\alpha_i d$ satisfying $\bida_i + \sum_{j = 1}
^n \bida_j c_j = 0$.

We define the $\beta_i$ above so that $\beta_i$ is the expansion of $f_i$ when viewed as a polynomial in $k[f_1,\ldots, f_{m+1}]$. Furthermore, the relations given in Equation ~\eqref{eqn:pm-relation-degree}
generate all relations, since they allow us to reduce any $\prod_{j =
1}^n f_j^{r_i}$ to a canonical form, with $r_i = -\lfloor d \alpha_i\rfloor$ whenever $i  > m + 1$.

For the remainder of the proof, fix $i \in \{1,\ldots, n\}$. 
To complete the proof, it suffices to bound the degree of generation of the relations of the form $(z_i - \beta_i) \prod_{j=1}^n z_j^{c_j}$, by 
Equation ~\eqref{eqn:pm-relation-degree}.  We first associate $(z_i - \beta_i)\prod_{j=1}^n z_j^{c_j}$ with $(d, c_1, \ldots, c_n) \in \Sigma,$ as defined in Equation ~\eqref{eqn:Sigma-def}.  
Let $\Sigma_i \subseteq \mathbb{Z}^{n+1}$ be the set of points of the form $(d, r_1, \ldots, r_n)$ satisfying $r_j \ge -d \alpha_j$ for all $j$ and $\sum_{j=1}^n r_j a_j = -a_i$.  Let 
\begin{align*}
	\delta_i = \left(\frac{a_i}{\deg(D)}, -\frac{\alpha_1 a_1}{\deg(D)}, \ldots, - \frac{\alpha_n a_n}{\deg(D)} \right).
\end{align*}
Then we see $\Sigma_i \subseteq \delta_i + \newspan_{\br_{\geq 0}}(e_1, \ldots, e_n)$ with $e_i$ as defined in Equation \ref{defn:e-i-proj}.
Therefore, we can write any element of $\Sigma_i$ as
\[
	\delta_i + \sum_{j=1}^n r_j e_j
\]
where $r_j \in \mathbb{R}$ for each $j$.

@@

\begin{thm}
\label{thm:proj-generators-relations}
Let $D = \sum_{i=1}^n \alpha_i D_i \in \bq \otimes \di \bp^m$. Let $f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that $D_i = V(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degree at most 
\[
	\sum_{i=1}^n \ell_i a_i
\]
with relations generated in degree at most
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{thm}
\begin{proof}
The bound on degree of generation is precisely the content of Lemma ~\ref{lem:proj-generators}. It only remains to bound the degree of relations.
By ~\ref{lem:bound-ker-chi}, $\ker(\chi)$ is generated in degrees up to $2\sum_{i=1}^n \ell_i a_i$ and by Lemma ~\ref{lem:proj-relations-psi}, $\ker(\psi),$ is generated in degrees up to $\frac{\bida_i}{\deg(D)} + \sum_{i=1}^n \ell_i a_i.$ Consequently, Lemma ~\ref{lem:composite-map} implies that $\ker(\phi)$ is generated in degrees less than
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{proof}

@@

By Proposition ~\ref{prop:cone-generation}, in order to determining a spanning set for $(R_D)_d,$ it suffices to find the extremal rays of $\Sigma.$
To do this, we extend the method of O'Dorney \cite[Theorem 8]{dorney:canonical}. 
We first consider the subcone $\Sigma_1 \subset \Sigma$
given by
\begin{align*}
	\Sigma_1 = \{ (d, c_1, \ldots, c_n) \in \bz_{\geq 0}^{n+1} : c_i \ge - d \alpha_i \text{ for all }i \text{ and }\sum_{i=1}^n c_i (\bida_i+\bidb_i) = 0 \}.	
\end{align*}
Then, $\Sigma_1$
has extremal rays given by

@@

For $i \in \Tp(D)$ and $j \in \Tm(D)$, define
\[
	\epsilon_{i, j} := \left(\prod_{k \ne i, j} f_k^{-\alpha_k \ell_{i,j} (\bida_i \bidb_
	j - \bida_j \bidb_i)}) {f_i}^{s_1} {f_j}^{s_2} \in H^0(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j \bidb_i) D \right)
\]

@@

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degrees less than $\sigma$ (as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through the semigroup ring 
\[
	\bk[\Sigma] = \left \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i
	\right \rangle.
\]
by
\[
\begin{tikzcd}
	\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
	x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

>>>>

\begin{lem}
\label{lem:proj-relations-psi}
$\ker(\psi)$ is generated in degrees up to
\begin{align}
\label{eqn:proj-relation-degree}
	\frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} +  \sum_{i=1}^n \ell_i a_i.
\end{align}

@@

\noindent
for all $d \in \mathbb{N}$ and $c_i \ge -\alpha_i d$ satisfying
$\bida_i + \sum_{j = 1}^n \bida_j c_j = 0$.

We define the $\beta_i$ above so that $\beta_i$ is the expansion of
$f_i$ when viewed as a polynomial in $k[f_1,\ldots, f_{m + 1}]$.
Furthermore, the relations given in Equation
~\eqref{eqn:proj-relation-degree} generate all relations, since they
allow us to reduce any $ \prod_{j = 1}^n f_j^{r_i}$ to a canonical
form, with $r_i = -\lfloor  d \alpha_i\rfloor$ whenever $i  > m + 1$.

For the remainder of the proof, fix $i \in \{1,\ldots, n\}$. To
complete the proof, it suffices to bound the degree of generation
of the relations of the form $(z_i - \beta_i) \prod_{j = 1}^n
z_j^{c_j}$, by Equation ~\eqref{eqn:proj-relation-degree}. We first
associate $(z_i - \beta_i)\prod_{j=1}^n z_j^{c_j}$ with $(d, c_1,
\ldots, c_n) \in \Sigma,$ as defined in Equation
~\eqref{eqn:Sigma-def}. Let $\Sigma_i \subseteq \mathbb{Z}^{n + 1}$
be the set of points of the form $(d, r_1, \ldots, r_n)$ satisfying
$r_j \ge -d \alpha_j$ for all $j$ and $\sum_{j=1}^n r_j a_j = -a_i$.
Let
\begin{align*}
	\delta_i = \left(\frac{a_i}{\deg(D)}, -\frac{\alpha_1 a_1}{\deg(D)},
	\ldots, - \frac{\alpha_n a_n}{\deg(D)} \right).
\end{align*}

\noindent
Then we see $\Sigma_i \subseteq \delta_i + \newspan_{\br_{\geq 0}}
(e_1, \ldots, e_n)$ with $e_i$ as defined in Equation
~\ref{defn:e-i-proj}. Therefore, we can write any element of
$\Sigma_i$ as
\[
	\delta_i + \sum_{j=1}^n r_j e_j
\]
where $r_j \in \mathbb{R}$ for each $j$.

@@

\begin{thm}
\label{thm:proj-generators-relations}
Let $D = \sum_{i=1}^n \alpha_i D_i \in \bq \otimes \di \bp^m$. Let
$f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that
$D_i = V(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degree at most 
\[
	\sum_{i=1}^n \ell_i a_i
\]

\noindent
with relations generated in degree at most
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}
	(	\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{thm}

\begin{proof}
The bound on degree of generation is precisely the content of Lemma 
~\ref{lem:proj-generators}. It only remains to bound the degree of 
relations.

By ~\ref{lem:bound-ker-chi}, $\ker(\chi)$ is generated in degrees 
up to $2\sum_{i=1}^n \ell_i a_i$ and by Lemma ~\ref{lem:proj-
relations-psi}, $\ker(\psi),$ is generated in degrees up to $\frac{
\bida_i}{\deg(D)} + \sum_{i=1}^n \ell_i a_i.$ Consequently, Lemma ~
\ref{lem:composite-map} implies that $\ker(\phi)$ is generated in 
degrees less than
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}
	(\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{proof}

@@

By Proposition ~\ref{prop:cone-generation}, in order to determining a spanning set for $(R_D)_d,$ it suffices to find the extremal rays of $\Sigma.$
To do this, we extend the method of O'Dorney \cite[Theorem 8]{dorney:canonical}. 
We first consider the sub-cone $\Sigma_1 \subset \Sigma$
given by
\begin{align*}
	\Sigma_1 = \{ (d, c_1, \ldots, c_n) \in \bz_{\geq 0}^{n+1} : c_i \ge - d \alpha_i \text{ for all }i \text{ and }\sum_{i=1}^n c_i (\bida_i+\bidb_i) = 0 \}.	
\end{align*}
Then, $\Sigma_1$
has extremal rays given by

@@

\noindent
For $i \in \Tp(D)$ and $j \in \Tm(D)$, define
\[
	\epsilon_{i, j} := \left(\prod_{k \ne i, j} f_k^{-\alpha_k \ell_{i,j} (\bida_i \bidb_
	j - \bida_j \bidb_i)}) {f_i}^{s_1} {f_j}^{s_2} \in H^0(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j \bidb_i) D \right)
\]

@@

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degrees less than $\sigma$ (as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through the semigroup ring 
\[
	\bk[\Sigma] = \left \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i
	\right \rangle.
\]

\noindent
by
\[
\begin{tikzcd}
	\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
	x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]
